\Rpackage{MSnbase}: labelled MS2 data pre-processing, visualisation and quantification.

### 2011-05-11 inst/doc/MSnbase-demo/MSnbase-demo.Rnw
new makeImpuritiesMatrix, itraqdata and writeMgfData
git-svn-id: https://hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MSnbase@55442 bc3139a8-67e5-0310-9ffc-ced21a209358

\Rpackage{MSnbase} imports raw MS data stored in \texttt{mzXML} 
format \citep{Pedrioli2004} with the function \Rfunction{readMzXMLData} function. 
More formats, notably \texttt{mzML} \citep{Martens2010}, will be added at a later stage. 
Either MS1 or MS2 spectra can be loaded at a time by setting the \texttt{msLevel} parameter accordingly.

<<MSnExp,cache=FALSE,echo=TRUE>>=
library("MSnbase")
filename <- dir(system.file(package="MSnbase",dir="extdata"),
                full.name=TRUE,pattern="mzXML$")
print(filename)
raw.experiment <- readMzXMLData(filename, msLevel = 2, verbose=FALSE)
raw.experiment
@

<<experiment-size,echo=FALSE,cache=FALSE>>=
sz <- sum(sapply(assayData(raw.experiment),object.size)) + 
  object.size(raw.experiment)
sz <- round(sz/(1024^2),2)
@

@@

The raw data is composed of the \Sexpr{length(raw.experiment)} MS spectra. 
The spectra are named individually 
(\Sexpr{paste(paste(head(featureNames(raw.experiment)),collapse=", "),", ...",sep="")})
and stored in a \Robject{environment}. They can be accessed individually with 
\texttt{raw.experiment[["X10"]]} or as a list with \texttt{spectra(raw.experiment)}.
As we have loaded our experiment specifying \texttt{msLevel=2}, 
the spectra will all be of level 2 (or higher, if available).

<<Spectrum,cache=FALSE,echo=TRUE>>=
sp <- raw.experiment[["X43"]]
sp
@

@@

<<accessors,cache=FALSE,echo=TRUE>>=
peaksCount(sp)
head(peaksCount(raw.experiment))
rtime(sp)
head(rtime(raw.experiment))
@

@@

It is also possible to plot all spectra of an experiment (figure \ref{fig:msnexp-plot}). 
Lets start by extracting all spectra that have the same precursor MZ value than \Robject{sp} 
into a separate experiment, using the \Rfunction{extractPrecSpectra} method. 
This method takes an \Robject{MSnExp} experiment and a precursor MZ value as parameters.

<<extractPrecSpec,echo=T,cache=F>>=
exp2 <- extractPrecSpectra(raw.experiment,precursorMz(sp))
exp2
@

We see that \Sexpr{length(exp2)} spectra have the same MZ value. 
These can be visulaised together by plotting the 
\Robject{MSnExp} object, as illustrated on figure \ref{fig:msnexp-plot}.

\begin{figure}[!ht]
<<msnexp-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=6>>=
plot(exp2,reporters=iTRAQ4,full=FALSE)
@ 
\caption{Experiment-wide raw MS2 spectra.}
\label{fig:msnexp-plot}
\end{figure}

@@

There are several methods implemented to perform basic data manipulation. 
Low intensity peaks can be set to 0 with the \Rfunction{removePeaks} method 
from spectra or whole experiments. The intensity threshold below which peaks are removed 
is defined by the \Robject{t} parameter. \Robject{t} can be specified directly 
as a numeric. The default value is the character \texttt{"min"}, that will remove 
all peaks equal to the lowest non null intensity in any spectrum. 
We observe the effect of the \Rfunction{removePeaks} method by comparing 
total ion count (i.e. the total intensity in a spectrum) with the \Rfunction{tic} method
before (object \Robject{sp}) and after (object \Robject{experiment}). 
Figure \ref{fig:spectrum-plot2} illustrated spectrum \texttt{X43} after low intensity 
peaks removal.

<<removePeaks,echo=TRUE,cache=FALSE>>=
experiment <- removePeaks(raw.experiment,t=3,verbose=FALSE)
## total ion current
tic(sp)
tic(experiment[["X43"]])
@

\begin{figure}[!ht]
<<spectrum-plot2,fig=T,echo=T,tikz=T,external=F,width=5,height=5>>=
plot(experiment[["X43"]],reporters=iTRAQ4,full=TRUE) 
@ 
\caption{Same spectrum as on figure \ref{fig:spectrum-plot}, 
  after setting peaks <= 3 to 0.}
\label{fig:spectrum-plot2}
\end{figure}

@@

<<clean,echo=TRUE,cache=FALSE>>=
## number of peaks
peaksCount(sp)
peaksCount(experiment[["X43"]])
experiment <- clean(experiment,verbose=FALSE)
peaksCount(experiment[["X43"]])
@

@@

\begin{figure}[p]
  \begin{center}
<<preproc-plot,fig=T,echo=F,tikz=T,external=F,width=3,height=6>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
print(p1, vp=vplayout(1,1))
print(p2, vp=vplayout(2,1))
print(p3, vp=vplayout(3,1))
@ 
\caption{This figure illustrated the effect or the \Rfunction{removePeaks} 
  and \Rfunction{clean} methods. The left-most spectrum displays two peaks, 
  of max height 3 and 7 respectively. The middle spectrum shows the result 
  of calling \Rfunction{removePeaks} with argument \Robject{t=3}, which sets 
  all data points of the first peak, whose maximum height is smaller or equal 
  to \Robject{t} to 0. The second peak is unaffected. Calling \Rfunction{clean} 
  after \Rfunction{removePeaks} effectly deletes successive 0 intensities from 
  the spectrum, as shown on the right plot. }
\label{fig:preproc}
  \end{center}
\end{figure}

@@

<<trimMz,echo=TRUE,cache=FALSE>>=
range(mz(sp))
experiment <- trimMz(experiment,mzlim=c(112,120))
range(mz(experiment[["X43"]]))
experiment
@

As can be seen above, all processing performed on the experiment is recorded 
and displayed as integral part of the experiment object.

@@

@@

<<makeGroups1,echo=FALSE,cache=FALSE>>=
gb <- as.factor(fData(qnt)$precursor.mz)
@

\Rfunction{combineFeatures}'s first argument, \Robject{object}, is an instance of class \Robject{MSnSet}, 
as has been created in the section \ref{sec:quant} for instance. 
The second argument, \Robject{groupBy}, is a \Robject{factor} than has as many elements as 
there are features in the \Robject{MSnSet} \Robject{object} argument. The features corresponding 
to the \Robject{groupBy} levels will be aggregated so that the resulting \Robject{MSnSet} output 
will have \Rfunction{length(levels(groupBy))} features. Here, we will combine individual 
MSMS spectra based on their precursor mass-to-charge ratio. This information has been saved 
in the \Robject{MSnSet}'s \Robject{featureData} slot during the quantitation (see section \ref{sec:quant}). 
As shown below, this will result in \Sexpr{length(table(gb))} new aggregated features.

<<makeGroups2,echo=TRUE,cache=FALSE>>=
names(fData(qnt))
head(fData(qnt)$precursor.mz)
gb <- as.factor(fData(qnt)$precursor.mz)
table(gb)
@

@@

Other intended ways to define aggregation groups is to import 
identification data into R and collate it to the \Robject{MSnSet} 
feature data, and subsequently use peptide sequences or protein assignments 
to combine features.

@@

<<incompdiss,echo=TRUE,cache=FALSE,keep.source=T>>=
iTRAQ5
incompdiss <- quantify(raw.experiment,
                       method="trap",
                       reporters=iTRAQ5,
                       strict=FALSE,
                       verbose=FALSE)
head(exprs(incompdiss))
@

>>>>

\Rpackage{MSnbase} imports raw MS data stored in \texttt{mzXML} 
format \citep{Pedrioli2004} with the function \Rfunction{readMzXMLData} function. 
See \Rfunction{?readMzXMLData} for more details.
More formats, notably \texttt{mzML} \citep{Martens2010}, will be added at a later stage. 
Either MS1 or MS2 spectra can be loaded at a time by setting the \texttt{msLevel} parameter accordingly.
In this document, we will use a slightly bigger dataset, \Robject{itraqdata}, provided with 
\Rpackage{MSnbase}. It includes feature metadata, accessible with the \Rfunction{fData} method. 
The metadata includes identification data for the \Sexpr{length(itraqdata)} MS2 spectra.

<<MSnExp,cache=FALSE,echo=TRUE>>=
library("MSnbase")
itraqdata
head(fData(itraqdata))
@

<<experiment-size,echo=FALSE,cache=FALSE>>=
sz <- sum(sapply(assayData(itraqdata),object.size)) + 
  object.size(itraqdata)
sz <- round(sz/(1024^2),2)
@

@@

The raw data is composed of the \Sexpr{length(itraqdata)} MS spectra. 
The spectra are named individually 
(\Sexpr{paste(paste(head(featureNames(itraqdata)),collapse=", "),", ...",sep="")})
and stored in a \Robject{environment}. They can be accessed individually with 
\texttt{itraqdata[["X10"]]} or as a list with \texttt{spectra(itraqdata)}.
As we have loaded our experiment specifying \texttt{msLevel=2}, 
the spectra will all be of level 2 (or higher, if available).

<<Spectrum,cache=FALSE,echo=TRUE>>=
sp <- itraqdata[["X1"]]
sp
@

@@

<<accessors,cache=FALSE,echo=TRUE>>=
peaksCount(sp)
head(peaksCount(itraqdata))
rtime(sp)
head(rtime(itraqdata))
@

@@

<<bsasel,echo=FALSE>>=
bsasel <- fData(itraqdata)$ProteinAccession=="BSA"
@

It is also possible to plot all spectra of an experiment (figure \ref{fig:msnexp-plot}). 
Lets start by subsetting the \Robject{itraqdata} experiment using the protein accession 
numbers included in the feature metadata, and keep the \Sexpr{sum(bsasel)} from 
the \textit{BSA} protein.

<<subset,echo=TRUE>>=
sel <- fData(itraqdata)$ProteinAccession=="BSA"
bsa <- itraqdata[sel]
bsa
as.character(fData(bsa)$ProteinAccession)
@

%% Lets start by extracting all spectra that have the same precursor MZ value than \Robject{sp} 
%% into a separate experiment, using the \Rfunction{extractPrecSpectra} method. 
%% This method takes an \Robject{MSnExp} experiment and a precursor MZ value as parameters.
%% <<extractPrecSpec,echo=T,cache=F>>=
%% exp2 <- extractPrecSpectra(raw.experiment,precursorMz(sp))
%% exp2
%% @

These can then be visualised together by plotting the 
\Robject{MSnExp} object, as illustrated on figure \ref{fig:msnexp-plot}.

\begin{figure}[!ht]
<<msnexp-plot,fig=T,echo=T,tikz=T,external=F,width=5,height=6>>=
plot(bsa,reporters=iTRAQ4,full=FALSE)
@ 
\caption{Experiment-wide raw MS2 spectra. The y-axes of the individual 
  spectra are automatically rescaled to the same range. }
\label{fig:msnexp-plot}
\end{figure}

@@

There are several methods implemented to perform basic data manipulation. 
Low intensity peaks can be set to 0 with the \Rfunction{removePeaks} method 
from spectra or whole experiments. The intensity threshold below which peaks are removed 
is defined by the \Robject{t} parameter. \Robject{t} can be specified directly 
as a numeric. The default value is the character \texttt{"min"}, that will remove 
all peaks equal to the lowest non null intensity in any spectrum. 
We observe the effect of the \Rfunction{removePeaks} method by comparing 
total ion count (i.e. the total intensity in a spectrum) with the \Rfunction{tic} method
before (object \Robject{itraqdata}) and after (object \Robject{experiment}) for 
spectrum \texttt{X55}.

<<removePeaks,echo=TRUE,cache=FALSE>>=
experiment <- removePeaks(itraqdata,t=400,verbose=FALSE)
## total ion current
tic(itraqdata[["X55"]])
tic(experiment[["X55"]])
@

\begin{figure}[!ht]
<<spectrum-plot2,fig=T,echo=F,tikz=T,external=F,width=5,height=2.5>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(1, 2)))
vplayout <- function(x, y)
  viewport(layout.pos.row = x, layout.pos.col = y)
p1 <- plot(itraqdata[["X55"]],full=TRUE) + theme_gray(5) 
p2 <- plot(experiment[["X55"]],full=TRUE) + theme_gray(5) 
print(p1,vp=vplayout(1,1))
print(p2,vp=vplayout(1,2))
@ 
\caption{Same spectrum before (left) and after setting peaks <= 400 to 0.}
\label{fig:spectrum-plot2}
\end{figure}

@@

<<clean,echo=TRUE,cache=FALSE>>=
## number of peaks
peaksCount(itraqdata[["X55"]])
peaksCount(experiment[["X55"]])
experiment <- clean(experiment,verbose=FALSE)
peaksCount(experiment[["X55"]])
@

@@

\begin{figure}[p]
  \begin{center}
<<preproc-plot,fig=T,echo=F,tikz=T,external=F,width=3,height=6>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(3, 1)))
print(p1, vp=vplayout(1,1))
print(p2, vp=vplayout(2,1))
print(p3, vp=vplayout(3,1))
@ 
\caption{This figure illustrated the effect or the \Rfunction{removePeaks} 
  and \Rfunction{clean} methods. The left-most spectrum displays two peaks, 
  of max height 3 and 7 respectively. The middle spectrum shows the result 
  of calling \Rfunction{removePeaks} with argument \Robject{t=3}, which sets 
  all data points of the first peak, whose maximum height is smaller or equal 
  to \Robject{t} to 0. The second peak is unaffected. Calling \Rfunction{clean} 
  after \Rfunction{removePeaks} effectly deletes successive 0 intensities from 
  the spectrum, as shown on the right plot. }
\label{fig:preproc}
  \end{center}
\end{figure}

@@

<<trimMz,echo=TRUE,cache=FALSE>>=
range(mz(itraqdata[["X55"]]))
experiment <- trimMz(experiment,mzlim=c(112,120))
range(mz(experiment[["X55"]]))
experiment
@ 
As can be seen above, all processing performed on the experiment is recorded 
and displayed as integral part of the experiment object.

@@

If no peak is detected for a reporter ion peak, the respective 
quantitation value is set to \texttt{NA}. In our case, there 
is \Sexpr{sum(is.na(exprs(qnt)))} such case in row 
\Sexpr{which(is.na(exprs(qnt))) %% nrow(qnt)}.
We will remove the offending line, as we do not expect any missing  peaks.

<<removeNa,echo=TRUE>>=
sum(is.na(exprs(qnt)))
whichRow <- which(is.na(exprs(qnt))) %% nrow(qnt)
qnt <- qnt[-whichRow,]
sum(is.na(exprs(qnt)))
@

@@

<<makeGroups1,echo=FALSE,cache=FALSE>>=
gb <- fData(qnt)$ProteinAccession
@

\Rfunction{combineFeatures}'s first argument, \Robject{object}, is an instance of class \Robject{MSnSet}, 
as has been created in the section \ref{sec:quant} for instance. 
The second argument, \Robject{groupBy}, is a \Robject{factor} than has as many elements as 
there are features in the \Robject{MSnSet} \Robject{object} argument. The features corresponding 
to the \Robject{groupBy} levels will be aggregated so that the resulting \Robject{MSnSet} output 
will have \Rfunction{length(levels(groupBy))} features. Here, we will combine individual 
MS2 spectra based on the protein they originate from. 
As shown below, this will result in \Sexpr{length(table(gb))} new aggregated features.

<<makeGroups2,echo=TRUE,cache=FALSE>>=
gb <- fData(qnt)$ProteinAccession
table(gb)
@

@@

%% Other intended ways to define aggregation groups is to import 
%% identification data into R and collate it to the \Robject{MSnSet} 
%% feature data, and subsequently use peptide sequences or protein assignments 
%% to combine features.

@@

<<incompdiss,echo=TRUE,cache=FALSE,keep.source=T>>=
iTRAQ5
incompdiss <- quantify(itraqdata,
                       method="trap",
                       reporters=iTRAQ5,
                       strict=FALSE,
                       verbose=FALSE)
head(exprs(incompdiss))
@
